---
title: 'Session 5: Spatial analysis'
author: "Lasse Meyer"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

Highly multiplexed imaging technologies acquire the spatial distributions 
of molecule abundances across tissue sections. As such, analyzing single cells
in their spatial tissue context is a key strength of these technologies.

A number of software packages such as
[squidpy](https://squidpy.readthedocs.io/en/stable/),
[Giotto Suite](https://giottosuite.com) and
[Seurat](https://satijalab.org/seurat/articles/spatial_vignette_2.html) have
been developed to analyse and visualize cells in their spatial context. The
following Session will highlight the use of
[imcRtools](https://bioconductor.org/packages/release/bioc/html/imcRtools.html)
and other Bioconductor tools tools to visualize and analyse single-cell data
obtained from highly multiplexed imaging technologies.

We will first read in the spatially-annotated single-cell data processed in the 
previous sections.

```{r spatial-load-data, message=FALSE}
library(SpatialExperiment)
spe <- readRDS("data/spe.rds")
```

# Overview

Spatial relationships take place at different biological length scales 
and range from more "global" formation of cellular neighborhoods to "local" 
cell-to-cell interactions. Methods can also be broadly divided into 
unsupervised and supervised spatial methods. 

Many approaches either compare the observed versus expected number of cells 
around a given cell type ("point pattern") or utilize interaction graphs 
("spatial object graphs") to estimate clustering or interaction frequencies
between cell types.

## Spatial graphs

The [steinbock](https://bodenmillergroup.github.io/steinbock/latest/cli/measurement/) 
framework allows the construction of these spatial graphs. During image 
processing, we have constructed a spatial graph by expanding the individual cell 
masks borders by 4 pixels. 

The `imcRtools` package further allows the *ad hoc* construction of spatial
graphs directly using a `SpatialExperiment` or `SingleCellExperiment` object
while considering the spatial location (here centroids) of individual cells. The
[buildSpatialGraph](https://bodenmillergroup.github.io/imcRtools/reference/buildSpatialGraph.html)
function allows constructing spatial graphs by detecting the k-nearest neighbors
in 2D (`knn`), by detecting all cells within a given distance to the center cell
(`expansion`) and by Delaunay triangulation (`delaunay`).

When constructing a knn graph, the number of neighbors (`k`) needs to be set and
(optionally) the maximum distance to consider (`max_dist`) can be specified.
When constructing a graph via expansion, the distance to expand (`threshold`)
needs to be provided. For graphs constructed via Delaunay triangulation,
the `max_dist` parameter can be set to avoid unusually large connections at the
edge of the image.

```{r build-spatial-graphs, message=FALSE}
library(imcRtools)
spe <- buildSpatialGraph(spe, img_id = "sample_id", type = "knn", k = 20)
spe <- buildSpatialGraph(spe, img_id = "sample_id", type = "expansion", threshold = 20)
spe <- buildSpatialGraph(spe, img_id = "sample_id", type = "delaunay", max_dist = 50)
```

The spatial graphs are stored in `colPair(spe, name)` slots. These slots store
`SelfHits` objects representing edge lists in which the first column indicates
the index of the "from" cell and the second column the index of the "to" cell.
Each edge list is newly constructed when subsetting the object.

```{r show-colPairNames}
colPairNames(spe)
```

Here, `colPair(spe, "neighborhood")` stores the spatial graph constructed by
`steinbock`, `colPair(spe, "knn_interaction_graph")` stores the knn spatial
graph, `colPair(spe, "expansion_interaction_graph")` stores the expansion graph
and `colPair(spe, "delaunay_interaction_graph")` stores the graph constructed by
Delaunay triangulation.

```{r show-colPairExample}
#steinbock expansion graph
colPair(spe, "neighborhood")
```

## Spatial visualization

In the following example, we select one image for visualization purposes.
For this we use the `plotSpatial` function from `imcRtools`. Here, each dot 
(node) represents a cell and edges are drawn between cells in close physical 
proximity as detected by `steinbock` or the `buildSpatialGraph` function. 
Nodes are variably colored based on the cell type and edges are colored in grey.

```{r spatial-viz-1, message=FALSE, fig.width=7, fig.height=7}
library(ggplot2)
library(viridis)

# steinbock interaction graph 
plotSpatial(spe[,spe$sample_id == "Patient3_001"], 
            node_color_by = "celltype", 
            img_id = "sample_id", 
            draw_edges = TRUE, 
            colPairName = "neighborhood", 
            nodes_first = FALSE, 
            edge_color_fix = "grey") + 
    scale_color_manual(values = metadata(spe)$color_vectors$celltype) +
    ggtitle("steinbock interaction graph")

# knn interaction graph 
plotSpatial(spe[,spe$sample_id == "Patient3_001"], 
            node_color_by = "celltype", 
            img_id = "sample_id", 
            draw_edges = TRUE, 
            colPairName = "knn_interaction_graph", 
            nodes_first = FALSE,
            edge_color_fix = "grey") + 
    scale_color_manual(values = metadata(spe)$color_vectors$celltype) +
    ggtitle("knn interaction graph")

# expansion interaction graph 
plotSpatial(spe[,spe$sample_id == "Patient3_001"], 
            node_color_by = "celltype", 
            img_id = "sample_id", 
            draw_edges = TRUE, 
            colPairName = "expansion_interaction_graph", 
            nodes_first = FALSE, 
            directed = FALSE,
            edge_color_fix = "grey") + 
    scale_color_manual(values = metadata(spe)$color_vectors$celltype) +
    ggtitle("expansion interaction graph")

# delaunay interaction graph 
plotSpatial(spe[,spe$sample_id == "Patient3_001"], 
            node_color_by = "celltype", 
            img_id = "sample_id", 
            draw_edges = TRUE, 
            colPairName = "delaunay_interaction_graph", 
            nodes_first = FALSE,
            edge_color_fix = "grey") + 
    scale_color_manual(values = metadata(spe)$color_vectors$celltype) +
    ggtitle("delaunay interaction graph")
```

Nodes can also be colored based on the cells' expression levels (e.g.,
E-cadherin expression) and their size can be adjusted (e.g., based on measured
cell area).

```{r spatial-viz-2, fig.width=7, fig.height=7}
plotSpatial(spe[,spe$sample_id == "Patient3_001"], 
            node_color_by = "Ecad", 
            assay_type = "exprs",
            img_id = "sample_id", 
            draw_edges = TRUE, 
            colPairName = "neighborhood", 
            nodes_first = FALSE, 
            node_size_by = "area", 
            directed = FALSE,
            edge_color_fix = "grey") + 
    scale_size_continuous(range = c(0.1, 2)) +
    ggtitle("E-cadherin expression on steinbock graph")
```

Finally, the `plotSpatial` function allows displaying all images at once. This
visualization can be useful to quickly detect larger structures of interest 
(e.g. large cohesive tumor regions and B/BnT cell aggregation)

```{r spatial-viz-3, fig.height=12, fig.width=12}
plotSpatial(spe, 
            node_color_by = "celltype", 
            img_id = "sample_id", 
            node_size_fix = 0.5) + 
    scale_color_manual(values = metadata(spe)$color_vectors$celltype)
```

For a full documentation on the `plotSpatial` function, please refer to
`?plotSpatial`.

# Unsupervised spatial analysis

In the following section, we will highlight more unsupervised ways for spatial 
analysis. First, we will focus on "global" cellular neighborhood detection based on 
the 1) celltype neighborhood, 2) expression neighborhood and 3) point-process
based assessment. Second, we will analyze "local" cell-to-cell interactions using 
permutation-based testing.

## Cellular neighborhood analysis

The following section highlights the use of the `imcRtools` package to detect
cellular neighborhoods. This approach has been first proposed by [Goltsev et al. (2018)](https://www.sciencedirect.com/science/article/pii/S0092867418309048?via%3Dihub)
and [Schürch et al. (2020)](https://www.sciencedirect.com/science/article/pii/S0092867420308709?via%3Dihub) 
to group cells based on information contained in their direct
neighborhood. 

### Cell type-based neighborhood analysis

In the following code chunk we will use the 20-nearest neighbor graph as
constructed above to define the direct cellular neighborhood. The
[aggregateNeighbors](https://bodenmillergroup.github.io/imcRtools/reference/aggregateNeighbors.html)
function allows neighborhood aggregation in 2 different ways:

1.  For each cell the function computes the fraction of cells of a
    certain type (e.g., cell type) among its neighbors.
2.  For each cell it aggregates (e.g., mean) the expression counts
    across all neighboring cells (see next section). 

Based on these measures, cells can now be clustered into cellular
neighborhoods. We will first compute the fraction of the different cell
types among the 20-nearest neighbors and use kmeans clustering to group
cells into 6 cellular neighborhoods.

**Of note:** constructing a 20-nearest neighbor graph and clustering
using kmeans with `k=6` is only an example. It is recommended to perform 
a parameter sweep across different graph construction algorithms and different
parmaters `k` for kmeans clustering. Finding the best CN detection
settings is also subject to the question at hand.

```{r cn-analysis, fig.height=12, fig.width=12}
# By celltypes
spe <- aggregateNeighbors(spe, 
                          colPairName = "knn_interaction_graph", 
                          aggregate_by = "metadata", 
                          count_by = "celltype")

# View output
head(spe$aggregatedNeighbors)

# Perform clustering and visualization
set.seed(220705)
cn_1 <- kmeans(spe$aggregatedNeighbors, centers = 6)
spe$cn_celltypes <- as.factor(cn_1$cluster)

plotSpatial(spe, 
            node_color_by = "cn_celltypes", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    scale_color_brewer(palette = "Set3")
```

There are now different visualizations to examine the cell type composition 
of the detected cellular neighborhoods (CN). 

First, we can visualize the relative cell type abundance per CN (column-scaled).

```{r heatmap-1}
library(pheatmap)

for_plot <- prop.table(table(as.character(spe$cn_celltypes), spe$celltype), margin = 1)

pheatmap(for_plot, 
         color = colorRampPalette(c("dark blue", "white", "dark red"))(100), 
         scale = "column")
```

Second, we can observe per cell type the fraction of CN that they are distributed
across.

```{r heatmap-2}
for_plot <- prop.table(table(as.character(spe$cn_celltypes), spe$celltype), margin = 2)

pheatmap(for_plot, 
         color = viridis(100), 
         display_numbers = TRUE, 
         number_color = "white", 
         number_format = "%.2f")
```

Last, we also recommended to visualize some images to confirm the interpretation of
cellular neighborhoods. For this we can again use the `plotSpatial` function:

```{r spatial-cn,fig.height=8, fig.width=10}
plotSpatial(spe[,spe$sample_id == "Patient1_003"],
            img_id = "cn_celltypes", 
            node_color_by = "celltype", 
            node_size_fix = 0.7) +
    scale_color_manual(values = metadata(spe)$color_vectors$celltype)
```

CN 1 and CN 6 are mainly enriched for tumor cells with CN 6 forming the
tumor/stroma border. CN 3 is mainly enriched for B and BnT cells
indicating TLS-like structures. CN 5 is composed of aggregated plasma cells 
and most T cells.

### Expression-based 

A deviation of this approach working directly on the expression of
the neighboring cells instead of their phenotypes has been published by
[Kim et al. (2022)](https://www.nature.com/articles/s41592-022-01657-2). 

Here, we will now detect cellular neighborhoods by computing the mean
expression across the 20-nearest neighbors prior to kmeans clustering
(k=6).

```{r cn-expression, fig.height=12, fig.width=12}
# By expression
spe <- aggregateNeighbors(spe, 
                          colPairName = "knn_interaction_graph", 
                          aggregate_by = "expression", 
                          assay_type = "exprs",
                          subset_row = rowData(spe)$use_channel)

set.seed(220705)

cn_2 <- kmeans(spe$mean_aggregatedExpression, centers = 6)
spe$cn_expression <- as.factor(cn_2$cluster)

plotSpatial(spe, 
            node_color_by = "cn_expression", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    scale_color_brewer(palette = "Set3")
```

Also here, we can visualize the cell type composition of each cellular
neighborhood in different ways. 

```{r heatmap-1-exp}
for_plot <- prop.table(table(as.character(spe$cn_expression), spe$celltype), margin = 1)

pheatmap(for_plot, 
         color = colorRampPalette(c("dark blue", "white", "dark red"))(100), 
         scale = "column")
```

```{r heatmap-2-exp}
for_plot <- prop.table(table(as.character(spe$cn_expression), spe$celltype), margin = 2)

pheatmap(for_plot, 
         color = viridis(100), 
         display_numbers = TRUE, 
         number_color = "white", 
         number_format = "%.2f")
```

```{r spatial-cn-exp, fig.height=8, fig.width=10}
plotSpatial(spe[,spe$sample_id == "Patient1_003"],
            img_id = "cn_expression", 
            node_color_by = "celltype", 
            node_size_fix = 0.7) +
    scale_color_manual(values = metadata(spe)$color_vectors$celltype)
```

When clustering cells based on the mean expression within the direct
neighborhood, tumor cells are split across CN 6, CN 1 and CN 4 without
forming a clear tumor/stroma border. This result likely reflects
patient-to-patient differences in the expression of tumor markers.

CN 3 again contains B cells and BnT cells but also CD8 and undefined
cells, therefore it is less representative of potential TLS compared to CN 3 in
previous CN approach. CN detection based on mean marker expression is
therefore sensitive to staining/expression differences between samples
as well as lateral spillover due to imperfect segmentation. On the other hand,
it may be used to detect spatial patterns without the need to identify cell
types up front.

### Point-process-based

An alternative to the `aggregateNeighbors` function is provided by the
[lisaClust](https://bioconductor.org/packages/release/bioc/html/lisaClust.html)
Bioconductor package. In contrast to `imcRtools`, the
`lisaClust` package computes local indicators of spatial associations
(LISA) functions and clusters cells based on those. More precise, the
package summarizes L-functions from a Poisson point process model to
derive numeric vectors for each cell which can then again be clustered
using kmeans. All steps are supported by the `lisaClust` function which
can be applied to a `SingleCellExperiment` and `SpatialExperiment` object.

In the following example, we calculate the LISA curves within a 10µm, 20µm and
50µm neighborhood around each cell. Increasing these radii will lead to broader
and smoother spatial clusters. However, a number of parameter settings should be
tested to estimate the robustness of the results.

```{r lisaClust, fig.height=12, fig.width=12, message=FALSE}
library(lisaClust)

set.seed(22)
spe <- lisaClust(spe, 
                 k = 6,
                 Rs = c(10, 20, 50),
                 spatialCoords = c("Pos_X", "Pos_Y"),
                 cellType = "celltype",
                 regionName = "cn_point",
                 imageID = "sample_id")

plotSpatial(spe, 
            node_color_by = "cn_point", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    scale_color_brewer(palette = "Set3")
```

```{r heatmap-1-pp}
for_plot <- prop.table(table(as.character(spe$cn_point), spe$celltype), margin = 1)

pheatmap(for_plot, 
         color = colorRampPalette(c("dark blue", "white", "dark red"))(100), 
         scale = "column")
```

```{r heatmap-2-pp}
for_plot <- prop.table(table(as.character(spe$cn_point), spe$celltype), margin = 2)

pheatmap(for_plot, 
         color = viridis(100), 
         display_numbers = TRUE, 
         number_color = "white", 
         number_format = "%.2f")
```

```{r spatial-cn-pp, fig.height=8, fig.width=10}
# plotSpatial for CN
plotSpatial(spe[,spe$sample_id == "Patient1_003"],
            img_id = "cn_point", 
            node_color_by = "celltype", 
            node_size_fix = 0.7) +
    scale_color_manual(values = metadata(spe)$color_vectors$celltype)
```

In this case, CN 6 largely contains tumor cells but no CN is forming the
tumor/stroma border. CN 5 indicates TLS-like structures. CN 4 indicates T cell
subtypes and plasma cells are aggregated to CN 2.


## Cell-to-cell interaction analysis

The following section focuses on local cell-to-cell interactions. We describe 
how to observe and test the average number of interactions between 
cell labels (e.g. cell-types) within grouping levels (e.g. images) using permutation-
based testing. For the first descriptions of the testing approaches, 
please refer to [Shapiro et al., Nature Methods](https://www.nature.com/articles/nmeth.4391).

The `imcRtools` package exports the `countInteractions` and
`testInteractions` functions, which summarize all cell-cell interactions
per grouping level (e.g. image). As a result, a table is returned where
each row represents one of all possible cell-type/cell-type interactions
among all grouping levels. Missing entries or `NA`s indicate missing
cell-type labels for this grouping level. The next section gives details
on how interactions are summarized.

The [countInteractions](https://bodenmillergroup.github.io/imcRtools/reference/countInteractions.html) 
function counts the number of edges (interactions) between each set of unique cell 
labels per grouping level. Simplified, it counts for each cell of type A the number of
neighbors of type B. This count is averaged within each unique grouping
level (e.g. image) in four different ways:

1.  `method = "classic"`: The count is divided by the total number of
    cells of type A. The final count can be interpreted as "How many
    neighbors of type B does a cell of type A have on average?"

2.  `method = "conditional"`: Formerly named "histocat". 
    The count is divided by the number of cells of type A that have at 
    least one neighbor of type B. The final count can be interpreted as 
    "How many many neighbors of type B has a cell of type A on average, 
    given it has at least one neighbor of type B?"

3.  `method = "patch"`: For each cell, the count is binarized to 0 (less
    than `patch_size` neighbors of type B) or 1 (more or equal to
    `patch_size` neighbors of type B). The binarized counts are averaged
    across all cells of type A. The final count can be interpreted as
    "What fraction of cells of type A have at least a given number of
    neighbors of type B?"

4.  `method = "interaction"`: The count is divided by the total number of 
    interactions from cell type A. The final count can be interpreted as "The 
    fraction of interactions of cell type A that occur with cell type B."

The `countInteractions` returns a `DataFrame` containing the summarized
counts (`ct`) for all combinations of `from_label`, `to_label` and
`group_by`.

```{r countInteractions}
out <- countInteractions(spe,
                         group_by = "sample_id",
                         label = "celltype",
                         method = "interaction",
                         colPairName = "neighborhood")
head(out)
```

In the next instance, one can test if the obtained count is larger or
smaller compared to what is expected from a random distribution of cell
labels. For this, the [testInteractions](https://bodenmillergroup.github.io/imcRtools/reference/testInteractions.html) function permutes the cell labels `iter` times (default: 1000) and counts interactions 
as described above. This approach generates a distribution of the interaction count 
under a random distribution of cell labels. The observed interaction count is
compared against this Null distribution to derive empirical p-values:

`p_gt`: fraction of perturbations equal or greater than the observed
count

`p_lt`: fraction of perturbations equal or less than the observed count

`zscore`: (observed count - mean of permuted counts) / sd of permuted counts

Based on these empirical p-values, the `interaction` score (attraction
or avoidance), overall `p` value and significance by comparison to
`p_treshold` (`sig` and `sigval`) are derived. All results are returned
in form of a `DataFrame`. We set the seed within the `SerialParam` function 
for reproducibility. 

```{r testInteractions}
library(BiocParallel)

out <- testInteractions(spe,
                        group_by = "sample_id",
                        label = "celltype",
                        method = "interaction",
                        colPairName = "neighborhood",
                        BPPARAM = SerialParam(RNGseed = 22))
out
```

The returned `DataFrame` contains the test results per grouping level (in this case
the image ID, `group_by`), "from" cell type (`from_label`) and "to" cell type
(`to_label`). The `sigval` entry indicates if a pair of cell types is
significantly interacting (`sigval = 1`), if a pair of cell types is
significantly avoiding (`sigval = -1`) or if no significant interaction or
avoidance was detected.

These results can be visualized by computing the sum of the `sigval` entries
across all images:

```{r testInteractions-2, message=FALSE}
library(tidyverse)

out %>% as_tibble() %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE)) %>%
    ggplot() +
        geom_tile(aes(from_label, to_label, fill = sum_sigval)) +
        scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

In the plot above the red tiles indicate cell type pairs that were detected to 
significantly interact on a large number of images. On the other hand, blue
tiles show cell type pairs which tend to avoid each other on a large number 
of images. We can observe that tumor cells are mostly compartmentalized and are in
avoidance with other cell types. As expected, B cells interact with BnT cells; 
regulatory T cells interact with CD4+ T cells and CD8+ T cells. Most cell types
show self interactions indicating spatial clustering. 

We can also use the `plotInteractions` function to visualize interaction networks 
from an edge list, typically the output of the `testInteractions` function. 

Before visualization, the edge list can be filtered to focus on relevant 
cell-cell interactions, groupings, and significance. In the example below, 
we retain interactions which are significant in more than 70% of images 
and tag the edges as attraction or repulsion. We will also customize node and 
edge aesthetics directly using `plotInteractions` arguments. The plot can also 
be modified further using `ggplot2` syntax.

```{r plotInteractions-filter data, fig.height=8, fig.width=10}
library(dplyr)
library(ggraph)

out <- out %>% 
    as.data.frame() %>% 
    mutate(pair_id = paste(from_label, to_label, sep = "__"),
                                          is_sig = !is.na(sig))
summary_df <- out %>%
  group_by(pair_id) %>%
  summarise(
    sum_sigval = sum(sigval, na.rm = TRUE),
    total_count = sum(is_sig)
  ) %>% ungroup() %>%
  mutate(
    fraction_sig = sum_sigval / total_count,
    int_type = case_when(
      sum_sigval > 0 ~ "attraction",
      sum_sigval < 0 ~ "repulsion"
    )
  ) %>% filter(abs(fraction_sig) > 0.7)

subset_out <- out %>%
  semi_join(summary_df, by = "pair_id") %>%
  left_join(summary_df %>% select(pair_id, int_type), by = "pair_id")

edge_color <- setNames(c("red3", "blue3"),c("attraction", "repulsion"))

# Now we can visualize only these filtered edges
plotInteractions(subset_out,
                 spe,
                 group_by = "sample_id",
                 label = "celltype", 
                 node_color_by = "name",
                 node_size_by = "n_cells",
                 edge_color_by = "int_type",
                 edge_width_by = "ct",
                 graph_layout = "chord")+
    scale_color_manual(values = metadata(spe)$color_vectors$celltype)+
    scale_edge_colour_manual(values = edge_color)
```

We can again observe that most cell types are significantly self-interacting. Tumor cells
are very compartmentalized and void other cell types. Of note, B cells also interact with BnT cells, 
likely driven by TLS-like structures, and Tregs interact significantly with CD4+ T cells and CD8+ T cells.

# Supervised spatial analysis 

## Patch detection 

The previous section focused on detecting cellular neighborhoods and cell-to-cell
interactions in a rather unsupervised fashion. However, the `imcRtools` package 
also provides methods for detecting spatial compartments in a supervised fashion. The
[patchDetection](https://bodenmillergroup.github.io/imcRtools/reference/patchDetection.html)
function allows the detection of connected sets of similar cells as proposed by
[Hoch et al.](https://www.science.org/doi/10.1126/sciimmunol.abk1692). In the following example, we will use the `patchDetection` function to detect function to detect TLS-like structures in three steps:

1. Find connected sets of BnT cells (using the `steinbock` graph).  
2. Components which contain less than 10 cells are excluded.  
3. Expand the components by 2µm to construct a concave hull around the patch and
include cells within the patch.

```{r patchDetection, fig.height=12, fig.width=12}
spe <- patchDetection(spe, 
                      patch_cells = spe$celltype == "BnTcell",
                      img_id = "sample_id",
                      expand_by = 2,
                      min_patch_size = 10,
                      colPairName = "neighborhood", 
                      BPPARAM = SerialParam(RNGseed = 22))

# BnT cells
spe$celltype_BnT <- ifelse(spe$celltype %in% c("BnTcell"),"BnTcell", "Other")
plotSpatial(spe, 
            node_color_by = "celltype_BnT", 
            img_id = "sample_id", 
            node_size_fix = 0.5) + 
    scale_color_manual(values = metadata(spe)$color_vectors$celltype)


# Patch detection
plotSpatial(spe, 
            node_color_by = "patch_id", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    theme(legend.position = "none") 
```

Next, we can use the [distToCells](https://bodenmillergroup.github.io/imcRtools/reference/distToCells.html) 
to calculate the min, max, mean or median distance for each cell to a given cell type or class of
cells of interest, the function `distToCells` is available. Cells of interest
are defined via the `x_cells` parameter as `logical` (e.g. patches) and distances to for all
cells to those cells will be reported in a new column in the `colData` of the
`SingleCellExperiment`.

If the cells of interest form patches (many cells of the same type next to each
other) or similarly if a patch detection has previously been performed the
positive distances reflect the distances from cells outside of patches (or non cells of interest) 
to the patch cells (or cells of interest) and the negative distances reflect the distances from cellsinside the patches to the cells outside the patches. For `statistics` = "min" this corresponds to the distance from the cells outside to the closest point on the patch border and from the cells inside to the closest point on the patch border.

Here, we will calculate the minimum distance to TLS-like patches.

```{r minDistCells, fig.height=12, fig.width=12}
spe <- distToCells(spe, 
                   x_cells = !is.na(spe$patch_id), 
                   img_id = "sample_id")

plotSpatial(spe, 
            node_color_by = "distToCells", 
            img_id = "sample_id", 
            node_size_fix = 1) +
    scale_color_gradient2(low = "dark blue", mid = "white", high = "dark red")
```

Finally, we can observe the minimum distances to TLS-like patches in a cell type specific
manner.

```{r celltype-distance, warning = FALSE, message=FALSE}
library(ggridges)
ggplot(as.data.frame(colData(spe))) + 
    geom_density_ridges(aes(distToCells, celltype, fill = celltype)) +
    geom_vline(xintercept = 0, color = "black", linewidth = 1) +
    scale_fill_manual(values = metadata(spe)$color_vectors$celltype)+
    theme_classic()
```

## Differential interaction testing 

Lastly, one can be interested in detecting differential interactions between 
conditions of interest (e.g. images with TLS-like structures and ones without). 
Different statistical approaches have been proposed for differential interaction testing 
including ones based on point-pattern testing such as the [spicyR](https://bioconductor.org/packages/release/bioc/html/spicyR.html) 
R/Bioconductor package and permutation-based ones such as the ones featured in
`imcRtools`. 

Here, we will make use of our permutation-based testing results from the 
`testInteraction` function. For this, we will annotate the images with BnT cell 
patches as a surrogate for TLS-like structures and add this to the 
output `DataFrame` from our `testInteraction` results, called `out`. 

```{r TLS-anno}
# Annotate images with TLS-like patch
patch_images <- colData(spe) %>% as_tibble() %>%
    filter(!is.na(patch_id)) %>% 
    pull(sample_id) %>% 
    unique()

spe$TLS_like <- ifelse(spe$sample_id %in% patch_images, "TLS_like", "No_TLS_like")

# Add to cell-to-cell interaction findings
out$TLS <- spe$TLS_like[match(out$group_by, spe$sample_id)] 
```

We can then compare the `zscore`, observed count - mean of permuted counts) / sd of permuted counts, 
from permutaton testing for cell-type pairs of interest. Specifically, we will look at interactions
from BnT cells to themselves (homotypic interaction).

```{r TLS-diffinteraction, warning = FALSE}
out_filtered <- out %>% filter(pair_id %in% c("BnTcell__BnTcell"))

library(ggpubr)  
ggplot(out_filtered, aes(TLS, zscore, fill = TLS))+
    facet_wrap(~pair_id)+
    geom_boxplot(, outliers = FALSE)+
    scale_fill_brewer(palette = "Set2")+
    theme_classic()+
    stat_compare_means()
```

As expected, we observe trends for B cell interactions but tumor avoidance in images with 
TLS-like structures. This analysis can be expanded for other cell type pairs
and highlights the need for clear biological rationales for (supervised) spatial 
analysis. 

# Save objects

Finally, the generated data objects can be saved. 

```{r save-objects}
saveRDS(spe, "data/spe.rds")
```

# Tasks

**TASK 1 - Spatial visualization**: 
A) Visualize the cell type composition of CN3 across all images. 
B) Visualize CD20 expression for CN3 across all images. 

**TASK 2 - Cellular neighborhood analysis**: 
Compare single cell assignments between different cellular neighborhood approaches 
by visualizing the number of cells that are shared between different results 
in a pairwise fashion. Which cellular neighborhoods are conserved between methods?

**TASK 3 - Cell-to-cell interaction analysis**: 
Perform cell-to-cell interaction analysis with a different counting approach 
and compare your findings. Does the interpretation change? 

**TASK 4 - Patch detection and differential interaction testing**:
A) Compare the zscore of heterotypic interactions of BnT cells to 
B cells and tumor cells between images with and without TLS-like structures. 
What do you observe? 
B) Perform patch detection for tumor cells and visualize the results. 

# Further reading
Various other spatial analysis approaches are available. Find a selection of 
some approaches below for further reading:

**Cellular neighborhood analysis**
Non-deep learning methods: 
[Spatial-LDA](https://journals.sagepub.com/doi/10.1089/cmb.2019.0340)

Deep learning methods: 
[CellCharter](https://www.nature.com/articles/s41588-023-01588-4)
[CelloHood](https://www.biorxiv.org/content/10.1101/2025.11.10.687184v1)
[Space-GM](https://www.nature.com/articles/s41551-022-00951-w)
[CytoCommunity](https://www.nature.com/articles/s41592-023-02124-2)

**Cell-to-cell interaction analysis**
Point-pattern based methods: 
[spicyR](https://bioconductor.org/packages/release/bioc/html/spicyR.html) 
[Kontextual](https://www.cell.com/cell-reports-methods/fulltext/S2667-2375(25)00211-5)
[SpatioPath](https://www.nature.com/articles/s41467-025-57943-y) 

# Session info

```{r}
sessionInfo()
```
